<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Introduction to Kmisc}
\usepackage[utf8]{inputenc}
-->

<style>

/* 
 * Table styles taken from Twitter Bootstrap 2.3.2 
 * Other styles modified from knitr's base styling.
 */

body {
  display: block;
  margin: 0;
  background-color: #DDD;
  font-size: 14px;
  font-family: sans-serif;
}

td {
   font-family: sans-serif;
   background-color: white;
   font-size: 14px;
   margin: 8px;
}

.container {
  max-width: 700px;
  margin: 0 auto;
  margin-top: 14px;
  margin-bottom: 14px;
  background-color: white;
  padding: 40px;
  border-radius: 10px;
  box-shadow: 0 0 6px 6px #AAA;
}

tt, code, pre {
  white-space: pre-line;
   font-family: 'DejaVu Sans Mono', 'Droid Sans Mono', 'Lucida Console', Consolas, Monaco, monospace;
   margin: 0 auto;
   line-height: 20px;
}

p code,
li code {
  padding-left: 3px;
  padding-right: 3px;
  color: #d14;
  background-color: #f7f7f9;
  border: 1px solid #e1e1e8;
  border-radius: 2px;
}

h1, h2, h3, h4, h5, h6 {
  font-family: "Gill Sans";
  text-align: center;
}

h1 {
  font-size:2.2em;
}

h2 { 
  font-size:1.8em; 
}

h3 {
  font-size:1.4em; 
}

h4 { 
  font-size:1.0em; 
}

h5 { 
  font-size:0.9em; 
}

h6 { 
  font-size:0.8em; 
}

a:visited {
   color: rgb(50%, 0%, 50%);
}

pre code {
   display: block;
   padding: 0.5em;
   padding-left: 10px;
   padding-right: 10px;
   border-style: solid;
   border-width: 1px;
   border-radius: 4px;
   border-color: #ABC;
   box-shadow: 1px 1px 3px 0 #DDD;
   margin-bottom: 1em;
}

code.r, code.cpp {
   background-color: #F8F8F8;
   box-shadow: 1px 1px 3px 0 #AAA;
}

table, td, th {
  border: none;
}

blockquote {
   color:#666666;
   margin:0;
   padding-left: 1em;
   border-left: 0.5em #EEE solid;
}

hr {
   height: 0px;
   border-bottom: none;
   border-top-width: thin;
   border-top-style: dotted;
   border-top-color: #999999;
}

@media print {
   * { 
      background: transparent !important; 
      color: black !important; 
      filter:none !important; 
      -ms-filter: none !important; 
   }

   body { 
      font-size:12pt; 
      max-width:100%; 
   }
       
   a, a:visited { 
      text-decoration: underline; 
   }

   hr { 
      visibility: hidden;
      page-break-before: always;
   }

   pre, blockquote { 
      page-break-inside: avoid; 
   }

   tr, img { 
      page-break-inside: avoid; 
   }

   img { 
      max-width: 100% !important; 
   }

   @page :left { 
      margin: 15mm 20mm 15mm 10mm; 
   }
     
   @page :right { 
      margin: 15mm 10mm 15mm 20mm; 
   }

   p, h2, h3 { 
      orphans: 3; widows: 3; 
   }

   h2, h3 { 
      page-break-after: avoid; 
   }
}

div.page img {
  display: block;
  margin: 0 auto;
}

img.centered {
  display: block;
  margin: 0 auto;
}

/* table styles borrowed from bootstrap */
table {
  max-width: 100%;
  background-color: transparent;
}

th {
  text-align: left;
}

.table {
  /* width: 100%; */
  min-width: 400px;
  margin-left: auto;
  margin-right: auto;
  margin-bottom: 20px;
}

.table thead > tr > th,
.table tbody > tr > th,
.table tfoot > tr > th,
.table thead > tr > td,
.table tbody > tr > td,
.table tfoot > tr > td {
  padding: 8px;
  line-height: 1.428571429;
  vertical-align: top;
  border-top: 1px solid #dddddd;
}

.table thead > tr > th {
  vertical-align: bottom;
}

.table caption + thead tr:first-child th,
.table colgroup + thead tr:first-child th,
.table thead:first-child tr:first-child th,
.table caption + thead tr:first-child td,
.table colgroup + thead tr:first-child td,
.table thead:first-child tr:first-child td {
  border-top: 0;
}

.table tbody + tbody {
  border-top: 2px solid #dddddd;
}

.table .table {
  background-color: #ffffff;
}

.table-condensed thead > tr > th,
.table-condensed tbody > tr > th,
.table-condensed tfoot > tr > th,
.table-condensed thead > tr > td,
.table-condensed tbody > tr > td,
.table-condensed tfoot > tr > td {
  padding: 5px;
}

.table-bordered {
  border: 1px solid #dddddd;
}

.table-bordered > thead > tr > th,
.table-bordered > tbody > tr > th,
.table-bordered > tfoot > tr > th,
.table-bordered > thead > tr > td,
.table-bordered > tbody > tr > td,
.table-bordered > tfoot > tr > td {
  border: 1px solid #dddddd;
}

.table-striped > tbody > tr:nth-child(odd) > td,
.table-striped > tbody > tr:nth-child(odd) > th {
  background-color: #f9f9f9;
}

.table-hover > tbody > tr:hover > td,
.table-hover > tbody > tr:hover > th {
  background-color: #f5f5f5;
}

table col[class^="col-"] {
  display: table-column;
  float: none;
}

table td[class^="col-"],
table th[class^="col-"] {
  display: table-cell;
  float: none;
}

iframe {
  width: 700px;
  height: 500px;
}

/* Custom controls for hiding different parts of a document */
div.bottom-controls {
  min-width: 120px;
  position: fixed;
  padding: 10px;
  right: 1%;
  bottom: 0px;
  background-color: white;
  box-shadow: 0 0 2px 2px #888;
  border-radius: 10px 10px 0 0;
}

.hide-me {
  background-color: #DDD;
  height: 6px;
  padding: 0 !important;
  font-size: 0;
}

.hide-me * {
  display: none;
}

</style>

Quick Introduction to Kmisc
=====

Kmisc introduces a bunch of utility functions to make the R coding experience
a bit easier. Some of the most useful functions in the package are shown here.

```{r, tidy=FALSE}
set.seed(123)
library(data.table)
library(Kmisc)
library(lattice)
library(grid)
library(Rcpp)
library(knitr)
dat <- data.frame( x=letters[1:4], y=1:4, z=LETTERS[1:4] )
```

`without`: This function is used to remove columns from a `list` / `data.frame`.

```{r, tidy=FALSE}
## let's remove columns 'x' and 'z' from dat.
dat[ !(names(dat) %in% c('x', 'z')) ]
## I always find that syntax awkward. Let's use Kmisc::without instead.
without( dat, x, z )
## finally, we can also index by $ (note: NOT done by partial matching):
without( dat, dat$x, dat$z )
## this could be handy for vectors with very long names, if using an IDE with
## auto-complete
```

`extract`: Extract vectors from a data.frame or list. Although there is already
a good subsetting syntax for lists and vectors, I wanted a complementary
function for `without`.

```{r tidy=FALSE}
extract( dat, x, y)
```

`without.re, extract.re`: Extract variables whose names don't match / do match
a regular expression pattern.
```{r tidy=FALSE}
re_extract( dat, "[xy]")
re_without( dat, "[xy]")
```

`swap`: Replace elements in a vector.

```{r tidy=FALSE}
tDat <- dat ## make a temporary copy of dat

## Replace some elements in tDat$y
tDat$y <- swap( tDat$y, from=c(2, 4), to=c(20, 40) )
cbind( dat$y, tDat$y )
```

`factor_to_char`, `char_to_factor`: A set of functions that recurse through
a list / data.frame and set all elements that are characters to factors,
and vice versa.

```{r tidy=FALSE}
bDat <- data.frame( x=rnorm(10), y=sample(letters,10), z=sample(letters,10) )
str( bDat )
str( factor_to_char(bDat) )
```

`dapply`: The `data.frame` version of the `l/sapply` series of functions.

Why have this function when `sapply` still does much the same? I always get
frustrated with the fact that either an `array` or a `list` is returned
by sapply, but never a `data.frame`.

```{r, tidy=FALSE}
dat <- data.frame( x = rnorm(100), y = rnorm(100), z = rnorm(100) )
dapply( dat, summary )
```

`kMerge`: Left joins, aka. `merge( all.x=TRUE, ... )` without any mangling 
of the order.

```{r, tidy=FALSE}
dat1 <- data.frame( id=5:1, x=c("a","a","b","b","b"), y=rnorm(5) )
dat2 <- data.frame( id=c(1, 2, 4), z=rnorm(3) )

## default merge changes id order
merge( dat1, dat2, by="id", all.x=TRUE )
## even the sort parameter can't save you
merge( dat1, dat2, by="id", all.x=TRUE, sort=TRUE )
# kMerge keeps it as is
kMerge( dat1, dat2, by="id" )
```

`in_interval`: A fast C implementation for determing which elements of a 
vector `x` lie within an interval `[lo, hi)`.

```{r tidy=FALSE}
x <- runif(10)*10; lo <- 5; hi <- 10
print( data.frame( x=x, between_5_and_10=in_interval(x, lo, hi) ) )
```

`stack_list`: Use this to stack data.frames in a list. This can be useful if
e.g. you've run some kind of bootstrap procedure and have all your results
stored in as a list of data.frames -- even `do.call( rbind, dfs )` can be slow.
The difference is even more prominent when used on very large lists.

This is partially deprecated by `data.table::rbindlist` now, which has a much faster C implementation.

```{r tidy=FALSE}
dfs <- replicate(1E3, 
  data.frame(x=rnorm(10), y=sample(letters,10), z=sample(LETTERS,10)),
  simplify=FALSE
  )
str( stack_list(dfs) )
system.time( stack_list(dfs) )
system.time( do.call(rbind, dfs) )
```

Fast String Operations
-----

For whatever reason, R is missing some nice builtin 'string' functions. I've
introduced a few functions, implemented in C for speed, for two common
string operations.

`str_rev`: Reverses a character vector; ie, a vector of strings.
`str_rev2` is there if you need to reverse a potentially unicode string.

```{r tidy=FALSE}
str_rev( c("ABC", "DEF", NA, paste(LETTERS, collapse="") ) )
str_rev2( c("はひふへほ", "abcdef") )
```

`str_slice`: Slices a vector of strings at consecutive indices `n`.
`str_slice2` exists for potentially unicode strings.
```{r tidy=FALSE}
str_slice( c("ABCDEF", "GHIJKL", "MNOP", "QR"), 2 )
str_slice2( "ハッピー", 2 )
```

`str_sort`: sort a string.
```{r, tidy=FALSE}
str_sort("asnoighewgypfuiweb")
```

File I/O
-----

Sometimes, you get really large data files that just aren't going to fit into
RAM. You really wish you could split them up in a structured way, transform
them in some way, and then put them back together. One might consider a more
'enterprise' edition of the split-apply-combine framework (Hadoop and friends),
but one dirty alternative is to use C++ to munge through a text file and pull out
things that we actually want.

`split_file`: This function splits a delimited file into multiple files, according to
unique entries in a chosen column.

`extract_rows_from_file`: From a delimited text file, extract only the rows for
which the entries in a particular column match some set of items that you
wish to keep.

C++ Function Generators
-----

These functions wrap around generated source code to produce useful, fast
functions very quickly.

`Rcpp_tapply_generator`: Generate fast `tapply` style functions from C++
code through Rcpp. See the example.

```{r, tidy=FALSE}
dat <- data.frame( y=rnorm(100), x=sample(letters[1:5], 100, TRUE) )
tMean <- Rcpp_tapply_generator("return mean(x);")
with( dat, tMean(y, x) )
with( dat, tapply(y, x, mean) )
```

`Rcpp_apply_generator`: An apply function generator tailored to 2D matrices.
However, your function definition must return a scalar value.

```{r, tidy=FALSE}
aMean <- Rcpp_apply_generator("return mean(x);")
mat <- matrix( rnorm(100), nrow=10 )
aMean(mat, 2)
apply(mat, 2, mean)
```

Faster Versions of Commonly Used R Functions
--------------------------------------------

`tapply_`: This function operates like `tapply` but works faster through a 
faster factor generating function, as well as an optimized split. Note that
it is however restricted to the (common) case of your value and grouping
variables being column vectors.

```{r, tidy=FALSE}
library(microbenchmark)
y <- rnorm(1000); x <- sample(letters[1:5], 1000, TRUE)
tapply(y, x, mean)
tapply_(y, x, mean)
microbenchmark( times=10,
  tapply(y, x, mean),
  tapply_(y, x, mean),
  tMean(y, x)
  )
```

`melt_`: This function operates like `reshape2:::melt`, but works almost
entirely through C and hence is much faster. 

```{r, tidy=FALSE}
dat <- data.frame(
  id=LETTERS[1:5],
  x1=rnorm(5),
  x2=rnorm(5),
  x3=rnorm(5)
  )
print(dat)
melt_(dat, id.vars="id")
```

`factor_`: A faster, simpler implementation of `factor` through Rcpp. This might
be useful in some rare cases where speed is essential.

```{r, tidy=FALSE}
lets <- sample(letters, 1E6, TRUE)
stopifnot( identical(
  factor_(lets),
  factor(lets)
  ) )
microbenchmark( times=5,
  factor_(lets),
  factor(lets)
  )
```

`html`: Custom HTML in an R Markdown document.

```{r results='asis', tidy=FALSE}
html(
  table( class="table table-bordered table-striped table-condensed table-hover", ## bootstrap classes
    tr(
      td("Apples"),
      td("Bananas")
      ),
    tr(
      td("20"),
      td("30")
      )
    )
  )
```

`anatomy`, `anat`: Like `str`, but much faster. It won't choke on very large `data.frame`s.

```{r, tidy=FALSE}
df <- data.table(x=1, y=2)
str(df)
anatomy(df)
```

`htmlTable`: A wrapper to `knitr::kable` that allows us to set tag attributes.
Note that we require chunk options `results='asis'`.

```{r, tidy=FALSE, results='asis'}
library(rCharts)
df <- data.frame(x=runif(1000), gp=1:1000)
htmlTable( head(df, n=5) )
```

