<!--
%\VignetteEngine{knitr::docco_linear}
%\VignetteIndexEntry{Introduction to Kmisc}
\usepackage[utf8]{inputenc}
-->

Introduction to Kmisc
=====================

Kmisc introduces a grab-bag of utility functions that should be useful to various
kinds of `useR`s. Some of the most useful functions in the package are demoed
in this vignette.

```{r, tidy=FALSE}
set.seed(123)
library(data.table)
library(Kmisc)
library(lattice)
library(grid)
library(Rcpp)
library(knitr)
library(rbenchmark)
dat <- data.frame( x=letters[1:4], y=1:4, z=LETTERS[1:4] )
opts_chunk$set(
  results="markup"
)
```

`without`: This function is used to remove columns from a `list` / `data.frame`.

```{r, tidy=FALSE}
## let's remove columns 'x' and 'z' from dat.
tryCatch( dat[ -c('x', 'z') ], error=function(e) print(e$message) )
## oh :(
dat[ !(names(dat) %in% c('x', 'z')) ]
## I always find that a bit awkward. Let's use Kmisc's without instead.
without(dat, x, z)
```

`extract`: Extract vectors from a data.frame or list. Although there is already
a good subsetting syntax for lists and vectors, I wanted a complementary
function for `without`.

```{r tidy=FALSE}
extract(dat, x, y)
```

`re_without, re_extract`: Extract variables whose names don't match / do match
a regular expression pattern.

```{r tidy=FALSE}
re_extract(dat, "[xy]")
re_without(dat, "[xy]")
```

`enumerate`: Like `lapply`, but accepts functions with two arguments,
specifying both the value and the index at the current iteration.

```{r tidy=FALSE}
l <- replicate(3, 1:3, simplify=FALSE)
print(l)
enumerate(l, function(x, i) x^i)
```

`swap`: Replace elements in a vector.

```{r tidy=FALSE}
tDat <- dat ## make a temporary copy of dat

## Replace some elements in tDat$y
tDat$y <- swap( tDat$y, from=c(2, 4), to=c(20, 40) )
cbind( dat$y, tDat$y )
```

`factor_to_char`, `char_to_factor`: A set of functions that recurse through
a list / data.frame and set all elements that are characters to factors,
and vice versa.

```{r tidy=FALSE}
bDat <- data.frame( x=rnorm(10), y=sample(letters,10), z=sample(letters,10) )
str( bDat )
str( factor_to_char(bDat) )
```

`dapply`: The `data.frame` version of the `l/sapply` series of functions.

Why have this function when `sapply` still does much the same? I always get
frustrated with the fact that either an `array` or a `list` is returned
by sapply, but never a `data.frame`.

```{r, tidy=FALSE}
dat <- data.frame( x = rnorm(100), y = rnorm(100), z = rnorm(100) )
dapply( dat, summary )
```

`kMerge`: Left joins, aka. `merge( all.x=TRUE, ... )` without any mangling
of the order.

```{r, tidy=FALSE}
dat1 <- data.frame( id=5:1, x=c("a","a","b","b","b"), y=rnorm(5) )
dat2 <- data.frame( id=c(1, 2, 4), z=rnorm(3) )

## default merge changes id order
merge( dat1, dat2, by="id", all.x=TRUE )
## even the sort parameter can't save you
merge( dat1, dat2, by="id", all.x=TRUE, sort=TRUE )
# kMerge keeps it as is
kMerge( dat1, dat2, by="id" )
```

`in_interval`: A fast C implementation for determing which elements of a
vector `x` lie within an interval `[lo, hi)`.

```{r tidy=FALSE}
x <- runif(10)*10; lo <- 5; hi <- 10
print( data.frame( x=x, between_5_and_10=in_interval(x, lo, hi) ) )
```

`stack_list`: Use this to stack data.frames in a list. This can be useful if
e.g. you've run some kind of bootstrap procedure and have all your results
stored in as a list of data.frames -- even `do.call( rbind, dfs )` can be slow.
The difference is even more prominent when used on very large lists.

This is partially deprecated by `data.table::rbindlist` now, which has a much
faster C implementation.

```{r tidy=FALSE}
dfs <- replicate(1E2,
  data.frame(x=rnorm(10), y=sample(letters,10), z=sample(LETTERS,10)),
  simplify=FALSE
)
str( stack_list(dfs) )
system.time( stack_list(dfs) )
system.time( do.call(rbind, dfs) )
system.time( data.table::rbindlist(dfs) )
```

Coercion Functions
------------------

`Kmisc` comes with a number of functions that offer faster conversions between
base R types. The names should be self explanatory. In cases where it's possible,
we also allow the user to modify objects in place.

```{r tidy=FALSE}
l <- replicate(5, rnorm(5), simplify=FALSE)
invisible(list2df(l, inplace=TRUE))
class(l)
## see also df2list, mat2df, df2mat
```


Fast String Operations
----------------------

R is missing some nice builtin 'string' functions. I've
introduced a few functions for common string operations.

`str_rev`: Reverses a character vector; ie, a vector of strings.
`str_rev2` is there if you need to reverse a potentially unicode string.

```{r tidy=FALSE}
str_rev( c("ABC", "DEF", NA, paste(LETTERS, collapse="") ) )
str_rev2( c("はひふへほ", "abcdef") )
```

`str_slice`: Slices a vector of strings at consecutive indices `n`.
`str_slice2` exists for potentially unicode strings.
```{r tidy=FALSE}
str_slice( c("ABCDEF", "GHIJKL", "MNOP", "QR"), 2 )
str_slice2( "ハッピー", 2 )
```

`str_sort`: sort a string.
```{r, tidy=FALSE}
str_sort("asnoighewgypfuiweb")
```

`str_collapse`: Collapse a string using `Rcpp` sugar; operates like
R's `paste(..., collapse="")`, but works much faster.
```{r, tidy=FALSE}
str_collapse( c("ABC", "DEF", "GHI") )
```


File I/O
--------

Sometimes, you get really large data files that just aren't going to fit into
RAM. You really wish you could split them up in a structured way, transform
them in some way, and then put them back together. One might consider a more
'enterprise' edition of the split-apply-combine framework (Hadoop and friends),
but one dirty alternative is to use C++ to munge through a text file and pull out
things that we actually want.

`split_file`: This function splits a delimited file into multiple files, according to
unique entries in a chosen column.

`extract_rows_from_file`: From a delimited text file, extract only the rows for
which the entries in a particular column match some set of items that you
wish to keep.

C++ Function Generators
-----------------------

Use these functions to generate C++ / Rcpp-backed functions for common R-style
operations.

`Rcpp_tapply_generator`: Generate fast `tapply` style functions from C++
code through Rcpp. See `?Rcpp_tapply_generator` for more information.

```{r, tidy=FALSE, eval=FALSE}
n <- 1E5
dat <- data.frame( y=rnorm(n), x=sample(letters[1:5], n, TRUE) )
tMean <- Rcpp_tapply_generator("return mean(x);")
with( dat, tMean(y, x) )
with( dat, tapply(y, x, mean) )
benchmark(
  Kmisc=with( dat, tMean(y, x) ),
  R=with( dat, tapply(y, x, mean) ),
  replications=5
)
```

`Rcpp_apply_generator`: An apply function generator tailored to 2D matrices.
However, your function definition must return a scalar value.

```{r, tidy=FALSE, eval=FALSE}
aMean <- Rcpp_apply_generator("return mean(x);")
mat <- matrix( rnorm(1E3), nrow=100 )
aMean(mat, 2)
apply(mat, 2, mean)
benchmark(
  Kmisc=aMean(mat, 2),
  R=apply(mat, 2, mean)
)
```

Faster Versions of Commonly Used R Functions
--------------------------------------------

`tapply_`: This function operates like `tapply` but works faster through a
faster factor generating function, as well as an optimized split. Note that
it is however restricted to the (common) case of your value and grouping
variables being column vectors.

```{r, tidy=FALSE}
y <- rnorm(1E5); x <- sample(letters[1:5], 1E5, TRUE)
tapply(y, x, mean)
tapply_(y, x, mean)
benchmark( replications=10,
  tapply(y, x, mean),
  tapply_(y, x, mean),
  tMean(y, x)
)
```

`melt_`: This function operates like `reshape2:::melt`, but works almost
entirely through C and hence is much faster.

```{r, tidy=FALSE}
dat <- data.frame(
  id=LETTERS[1:5],
  x1=rnorm(5),
  x2=rnorm(5),
  x3=rnorm(5)
)
print(dat)
melt_(dat, id.vars="id")
```

`factor_`: A faster, simpler implementation of `factor` through Rcpp. This might
be useful in some rare cases where speed is essential.

```{r, tidy=FALSE}
lets <- sample(letters, 1E6, TRUE)
stopifnot( identical(
  factor_(lets),
  factor(lets)
) )
benchmark( replications=5,
  factor_(lets),
  factor(lets)
)
```

`anatomy`, `anat`: Like `str`, but much faster. It won't choke on very large `data.frame`s.

```{r, tidy=FALSE}
df <- data.table(x=1, y=2)
str(df)
anatomy(df)
```

Miscellaneous Functions
-----------------------

`html`: Custom HTML in an R Markdown document.

```{r results='asis', tidy=FALSE}
html(
  table( class="table table-bordered table-striped table-condensed table-hover", ## bootstrap classes
    tr(
      td("Apples"),
      td("Bananas")
    ),
    tr(
      td("20"),
      td("30")
    )
  )
)
```
